Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite thus set comparison cannot be done by exhaustively
comparing the points of one with the other. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
query is expressed in quantifier free linear arithmetic, which is decidable.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

\subsection{Infering specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code, which are inserted automatically as comments. This helps support
maintenance of legacy code, and aids understanding of our
specification language. We illustrate inference using the
five-point stencil of \Cref{exm:checking}.

The preceding source code analysis converts the concrete syntax of
array subscripts into a set of subscript schemes in
$\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
into sets of index schemes.

%
%For each array variable $a \in \mathsf{dom}(U)$, the algorithm
%constructs a set of $n$-dimensional rectangles covering all contiguous
%groups of schemes in $U(a)$.
%
We explain the algorithm in two parts. First, we show how
\emph{contiguous} index schemes are coalesced into a smaller set of
index schemes while remaining in union normal form. Then, we explain
how this smaller unioned indexing schemes are converted to
specifications.

\subsubsection{Covering}
A covering of (possibly overlapping) intervals is calculated as a
value of \region{N} by coalescing \emph{contiguous} index schemes
until a fixed-point is reached.

\newcommand{\contig}[2]{\mathit{contig}(#1, #2)}
\begin{defn}[Contiguity]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{contiguous} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
A coalesced set of index schemes is computed by:%
%
\begin{align*}
  \mathit{coalesce}(\mathcal{S}) = \; & \;\;
  \{\,S \cup T \mid S \in \mathcal{S}, T \in \mathcal{S} \, \wedge \,
                                        \contig{S}{T} \}
  \;\;\; \cup \;\;\;
  \{\,S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
\end{align*}
%
Thus, $\mathit{coalesce}(\mathcal{S})$
comprises the union of the set of coalesced pairs of contiguous
schemes and the set of index schemes for which there is no other
contiguous scheme.
%
\begin{lemma}[Closure]
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesce}(\mathcal{S})$ is well-defined and is a set of
  index schemes.
\end{lemma}
\begin{proof}
  By \Cref{lem:vector-union}, the union of two index schemes is an
  index scheme if all components are equal except at most one
  component. Contiguity (\Cref{def:contiguity}) matches
  this precondition, thus the first part of the union in
  $\mathit{coalesce}$ produces a set of index schemes. The second part
  comprises a subset of the original index schemes.
\end{proof}
%
For our example, the fixed-pointed of $\mathit{coalesce}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesce}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesce}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesce}^3(\mathcal{S}_0)
\end{align*}
%
\todo{Prove that this is in $Region_N$}

@@

Now we convert coalesced indexing schemes into specifications. This happens in
three stages. First, we see if there is an all infinite indexing scheme. Second,
we determine if the indexing schemes are interval schemes so that they can be
represented exactly and if not how it can be altered into interval schemes to be
represented as approximations. Third, we split up the interval schemes so that
the region constants can express them. Finally, we convert the interval schemes
into a joins of meets of region constants.

>>>>

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite thus equality cannot be computed by exhaustively
comparing elements in each set. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
query is expressed in quantifier free linear arithmetic, which is decidable.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are
connected with no more than two \texttt{+} operations in 100\% of the cases.

\subsection{Inferring specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code which are inserted automatically as comments. This supports
maintenance of legacy code base, and aids understanding of our
specification language.

The preceding source code analysis converts the concrete syntax of
array subscripts into a sets of subscript schemes. Inference then has
two parts. First, ``adjacent'' index schemes are coalesced into a
smaller sets of index schemes, remaining in union normal
form. Secondly, the resulting union normal form is converted to the
specification syntax.

\subsubsection{Covering}
A covering of (possibly overlapping) intervals is calculated
by coalescing \emph{contiguous} index schemes
until a fixed-point is reached.

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
we generate the set of coalescing of $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This give a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

\newcommand{\finalSet}{\mathcal{S}_\omega}
Let the covering indexing scheme from the fixed point of
$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
syntax. This happens in four stages.

\begin{enumerate}[leftmargin=1.5em]
  \item We check whether $\finalSet{}$
  is the top indexing scheme $\bz{}^N$. This occurs if absolute indices
  occur in each of the dimensions in any of the array subscripts,
  leading to an unconstrained access pattern, which is not represented
  in our syntax, ending inference.

  \item Otherwise, we determine if the indexing schemes are interval
  schemes so that they can be represented exactly. It not, they are
  altered into interval schemes to be represented as approximations.

  \item We split the interval schemes so that the region constants can
  express them.

  \item Finally, we convert the interval schemes into a joins of meets
  of region constants.
\end{enumerate}
